---
title: "exercise11"
format: html
---

library packages

```{r}
library(dplyr)
library(tidyverse)
library(skimr)
library(ggplot2)
library(MuMIn)
```

load in data

```{r}
d <- read_tsv("https://raw.githubusercontent.com/difiore/ada-datasets/main/Mammal_lifehistories_v2.txt")

skim(d)
```

Categorical variables: order, family, genus, species.

Numeric variables: mass, gestation, newborn, weaning, wean mass, AFR, max. life, litter size, litters/year, refs.

### step 1

```{r}
d [d == -999] <- NA
```

### step 2

```{r}
d <- d %>% dplyr::select(-`litter size`, -refs)
```

### step 3

```{r}
d <- d %>% mutate(across(where(is.numeric), log))
```

### step 4

```{r}
m1 <- lm(`gestation(mo)`~ `mass(g)`, data = d, na.action = na.exclude)
d$relGest <- residuals(m1) 

m2 <- lm(`weaning(mo)`~ `mass(g)`, data = d, na.action = na.exclude)
d$relWean <- residuals(m2)

m3 <- lm(`AFR(mo)`~ `mass(g)`, data = d, na.action = na.exclude)
d$relAFR <- residuals(m3)

m4 <- lm(`max. life(mo)`~ `mass(g)`, data = d, na.action = na.exclude)
d$relLife <- residuals(m4)

m5 <- lm(`newborn(g)`~ `mass(g)`, data = d, na.action = na.exclude)
d$relNewbornMass <- residuals(m5)

m6 <- lm(`wean mass(g)`~ `mass(g)`, data = d, na.action = na.exclude)
d$relWeaningMass <- residuals(m6)
```

### step 5

```{r}
#relLife ~ Order
ggplot(d, aes(x = order, y = relLife)) +
  geom_boxplot() +
  labs(title = "Relative Max Lifespan by Order", y = "Residual Lifespan")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The order Primates has the highest residual lifespan.

```{r}
#relNewbornMass ~ Order
ggplot(d, aes(x = order, y = relNewbornMass)) +
  geom_boxplot() +
  labs(title = "Relative Newborn Mass by Order", y = "Residual Newborn Mass")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The order Macroscelidea has the highest residual newborn mass.

```{r}
#relWeaningMass ~ Order
ggplot(d, aes(x = order, y = relWeaningMass)) +
  geom_boxplot() +
  labs(title = "Relative Weaning Mass by Order", y = "Residual Weaning Mass")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

The order Perissodactyla has the have highest residual newborn mass.

### step 6

```{r}
# for max.life(mo)
d1 <- d %>% dplyr::select(`max. life(mo)`, `gestation(mo)`, `newborn(g)`,`weaning(mo)`, `wean mass(g)`, `litters/year`, `mass(g)`) %>% drop_na()

d1m_full <- lm(data = d1,
             `max. life(mo)` ~ `gestation(mo)`+ `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year`+ `mass(g)`, na.action = na.fail)
mods1 <- dredge(d1m_full)
mods1_top <- mods1 %>% filter(delta <= 4)
mods1_top
```

From the result, the best model is:

`max. life(mo)` \~ `gestation(mo)`+ `litters/year`+ `mass(g)` + `weaning(mo)`

5 models have a delta AICc of 4 or less.

Gestation(mo), litters/year, and weaning(mo) appeared in all top models.

```{r}
mods1avg <- summary(model.avg(mods1, subset = delta <= 4, fit = T))
mods1avg
plot(mods1avg)
```

```{r}
# for AFR(mo)
d2 <- d %>% dplyr::select(`AFR(mo)`, `gestation(mo)`, `newborn(g)`,`weaning(mo)`, `wean mass(g)`, `litters/year`, `mass(g)`) %>% drop_na()

d2m_full <- lm(data = d2,
             `AFR(mo)` ~ `gestation(mo)`+ `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year`+ `mass(g)`, na.action = na.fail)
mods2 <- dredge(d2m_full)
mods2_top <- mods2 %>% filter(delta <= 4)
mods2_top
```

From the result, the best model is:

`AFR(mo)` \~ `gestation(mo)` + `litter/year` + `mass(g)` + `weaning(mo)`

7 models have a delta AICc of 4 or less.

Gestation(mo), litters/year, weaning(mo) appeared in all top models.

```{r}
mods2avg <- summary(model.avg(mods2, subset = delta <= 4, fit = T))
mods2avg
plot(mods2avg)
```
